---
title: "Metagenome Assembly"
author: "Ryan Waln"
date: "2026-07-09"
output: html_document
---

## Introduction (everything will be in raw_data)

This file serves as a set of notes on the commands used for workshop 3 for Module 3 in MB5370 in order to assemble a metagenome of five bacterial species

Challenges include:
- Variable coverage: Some species are very abundant whereas others are quite rare. Assembling the rarer taxa is more difficult.
- Similar sequences: This community contains two Vibrio species. These are similar enough that they are difficult to separate in an assembly

The following steps are taken to assemble the metagenome:
- Examine Reads
- Assembly
- Basic assembly QC
- Binning
- Bin assessment

## Step 1: Examine Reads

### Create a subdirectory called raw_data in Marine_Genomics and symbolic link to the raw reads file, then check that files are correct with ls command
ln -s /pvol/data/metagenome_assembly/reads_R10_0.5.fastq.gz raw_reads.fastq.gz

### Examine the read length distribution in our input data (will claim broken pipe, proceed anyways)
zcat raw_reads.fastq.gz | head -n 40000 | bioawk -c fastx '{print $name,length($seq)}' > read_lengths.txt

```{r, explanation of code above}
## Unzips the raw data and sends the data to standard output
# zcat raw_data/raw_reads.fastq.gz \

## This is a pipe. It sends output from the previous command as input to the next command
#  | \
## Here we select only the top 40k lines in the file. This equates to 10k reads. It should be plenty to get some stats
#  head -n 40000 | \

## This is a small bioawk program. It prints the name of each sequence followed by its length
# bioawk -c fastx '{print $name,length($seq)}' \

## Send outputs to a file called read_lengths.txt
# > read_lengths.txt
```

### Visualize Read Lengths
```{r}
library(tidyverse)
library(here)

rl_data <- read_tsv(here::here("raw_data","read_lengths.txt"), col_names = c("id","length"), show_col_types = FALSE)

ggplot(rl_data, aes(x=length/1000)) + 
  geom_histogram(binwidth = 2) +
  xlab("Read Length Kb") + ylab("Number of Reads")
```
### Calculate N50 Statistic
### Step 1: Rank all reads from longest to shortest
### Step 2: Create a new column with the cumulative sum of read lengths (starting from longest read).
### Step 3: Find the read at which the cumulative sum surpasses 50% of the total read volume (ie total sum of all read lengths)
```{r}
total_len <- sum(rl_data$length)

rl_data %>% 
  arrange(desc(length)) %>% 
  mutate(cumsum = cumsum(length)) %>% 
  filter(cumsum>total_len/2) %>% 
  head(n=1)
# N50 of the individual reads is about 22Kb. We expect that the N50 of the assembly should be considerably greater than this.
```

## Step 2: Assembly 

### Use flye to assemble reads so create a directory, create a sheel script with provided code and save to raw_data/flye as run_flye.sh
### Run in slurm and ensure output is being generated
### Once finished, outputs from flye will be available in the directory mock (use bandage to visualize)
sbatch run_flye.sh
tail -f slurm-394.out
### Large Depth correlates with repeats or more of that species being present

### When flye is finished running need to compare the assembled contigs with reference genomes from the 5 species in our mock community. 
### Will acomplish by tabulating chromosome lengths and their corresponding GC content.
### Create a plain text file in raw_data/flye and post in provided code

### All the reference genomes for our mock community can be found at the path
ls /pvol/data/mg/refs/*.fna

### Test awk program by using it to print the GC content and contig lengths for all chromosomes in the files above. 
### Use cat to concatenate all the files together and then a pipe to send their contents to our bioawk program
cat /pvol/data/mg/refs/*.fna | bioawk -c fastx -f gc.awk

## Step 3: Basic Assembly QC

### Create a new directory called assembly_check in raw_data
### Then use your gc.awk script to print GC and lengths for the assembly
cat ../flye/mock/assembly.fasta | bioawk -c fastx -f ../flye/gc.awk
### Run again to send outputs to file below
cat ../flye/mock/assembly.fasta | bioawk -c fastx -f ../flye/gc.awk > assembly.gc.tsv

### We have GC content in assembly.gc.tsv and coverage depth in the assembly_info.txt file produced by flye. Let’s read those and join them together using R.
```{r}
gc <- read_tsv("assembly_check/assembly.gc.tsv",col_names = c("#seq_name","length","GC"))
info <- read_tsv("flye/mock/assembly_info.txt") 

```

